Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
minor bib & manuscript tweaks (notes)

The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{verbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{verbatim}
to represent the seven allowed state transitions.

@@

\subsection{Our group then}
 - Modifying Formatt took nearly a year
 - mutable data structures hard to repurpose
 - one enhancement deemed infeasible
 - C++ failures most commonly segfaults, memory errors, heap exhaustion

@@

In summary, with Haskell, we didn't have to master a completely different 
design and development style but we did have to wrap our heads around 
functional programming.

>>>>

The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.

@@

\subsection{Our group then}
 - reimplementing algorithms
  - we reimplemented and yet it was easy and pleasant
  - past reimplementation of a variant of Viterbi DP, in haste, -> 700 line C function
  - building on that reimplementation
 - new implementations (Matt)
  - Modifying Formatt took nearly a year
  - mutable data structures hard to repurpose
  - one enhancement deemed infeasible
 - C++ failures most commonly segfaults, memory errors, heap exhaustion

@@

In summary, with Haskell, we didn't have to master a completely different 
design and development style but we did have to wrap our heads around 
functional programming.\remark{Andrew, look at the comments below this
part of the manuscript. See what you can do with them.}
